High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-09-30 manuscript/Manuscript.tex

\begin{abstract}
New high-quality absolute paleointensity data from anorthosite xenoliths within the Beaver River diabase intrusions of the North American Midcontinent Rift (MCR) record an average virtual axial dipole moment (VADM) estimate of $\sim$80 ZAm$^2$ \textit{ca.} 1092 Ma. This cooling rate corrected paleointensity estimate from the MCR intrusive unit is similar to previous results from extrusive units that are dated to be \textit{ca.} 1108 Ma and \textit{ca.} 1085 Ma. Rock magnetic analyses and magnetic imaging reveal that the anorthosites can have abundant single-domain magnetic carriers that are ideal for paleointensity experiment, leading to a high success rate passing our selection criteria. Taken together, new and previously published paleointensity data from the Midcontinent Rift support that a geomagnetic field strength persisted through the Midcontinent Rift development and the data are inconsistent with the interpretation of a decay of Earth's geodynamo strength through the late Mesoproterozoic.

@@

@@

Non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained from Midcontinent Rift rock units. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength whereas \citeA{Kulakov2013a} used the high-temperature slope. Such non-ideal behavior were interpreted as failure by \siteA{Sprain2018a} who applied stricter paleointensity selection criteria and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. Despite the low success rate of paleointensity experiments, specimens that yield dominantly single-slope paleointensity results from the \textit{ca.} 1105 Ma Osler Volcanic Group, Mamainse Point volcanics \cite{Sprain2018a}, as well as the \textit{ca.} 1085 Ma Lake Shore Traps \cite{Kulakov2013a} suggest VADM values similar to that of the present-day \cite{Sprain2018a}. Taken together, these rock units that have been dated to span $\sim$25 myr during the development of the Midcontinent Rift magmatism \cite{Swanson-Hysell2019a} may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

Although it is typical of paleointensity experiments to suffer from low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested plagioclase crystals can protect magnetic minerals that exsolved within them against thermochemical alteration during paleointensity experiments. In this study, we present new paleointensity results from ``IZZI" absolute paleointensity experiments \cite{Yu2004a} on a suite of \textit{ca.} 1092 Ma bulk anorthosite xenoliths and their diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. The anorthosite lithology yielded exceptionally high success rate thanks to the plagioclase crystals shielding the magnetic minerals within. Thermal remanent magnetization (TRM) anisotropy test, full TRM acquisition test, together with magnetic imaging show that the bulk anorthosites from the Beaver Bay Complex have low TRM anisotropy, linear TRM acquisition property, and have dominant magnetic carriers interstitial to the plagioclase and are well shielded. With more single-slope, high quality paleointensity data, we seek to compare paleointensity records from the Midcontinent Rift intrusives with those from the extrusives; to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift; and to further enrich the reliable paleointensity records during the $\sim$25 myr of Midcontinent Rift magmatic activity.

@@

\begin{table}
\caption{Quality criteria used for accepting or rejecting paleointensity data. See the text for more detail.} 
\centering
\begin{tabular}{ccccccccc}
\hline
MAD (o) & Beta (per cent) & DANG (o) & FRAC& SCAT & GAP-MAX & NpTRM & N Arai\\
\hline
20 & 15 & 5 & 0.6 & TRUE & 0.6 & 2 & 4 \\
\hline

@@

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{MPMS.pdf}
\centering
\caption{\footnotesize{Low-temperature magnetic property measurement system (MPMS) experiment results. In the field-cooled (FC) experiments, the magnetization was measured upon warming following the specimen having cooled in an applied field of 2.5 T from 300 to 10 K. In the zero-field-cooled (ZFC) experiment, a low-temperature saturation isothermal remanence (LTSIRM) of 2.5 T was applied at 10 K after the specimen cooled in a (near-)zero field. In the room-temperature saturation isothermal remanence (RTSIRM) experiment, the sample was pulsed with a 2.5 T field at room temperature ($\sim$300 K) and then cooled to 10 K and warmed back to room temperature in a (near)zero field. Specimen AX6-1r is from anorthosite AX6 which passed our paleointensity selection. It has a well-defined Verwey transition $\sim$120 K. Specimens AX3-2r-ni and BD2-2r have prominent Verwey transition but the transition temperatures are suppressed below $\sim$120 K. }}
\label{fig:MPMS}
\end{figure}

@@

Anyways. anisotropy of full TRM experiment on accepted specs show low degree of bulk sample scale full TRM anisotropy, with a range of ---. Cooling rate correction based on \cite{Halgedahl1980a} suggest a lab over estimate of B ancient of 35 \%, assuming constant cooling rate in natural condition. The cooling rate corrected field is $\sim$40 $\mu$T which is similar to today's field strength at similar latitude.

\acknowledgments
Project research was supported by NSF CAREER grant EAR-1847277 to N.L.S.-H. and an Institute on Lake Superior Geology Student Research Fund grant to Y.Z. Permits for fieldwork and sampling from the Minnesota Department of Natural Resources are gratefully acknowledged. We thank James Pierce for assistance in the field.

We thank Jim Miller for providing us with the Duluth Complex anorthosite thin section.

We thank Dario Bilardello, Peat Solheid, Mike Jackson, Josh Feinberg, and Bruce Moskowitz for their tremendous help of instrumental operations, data interpretations, and research guidance. Y.Z. thanks the IRM for the U.S. Visiting Student Fellowship. Paleomagnetic data associated with this study are available within the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED) and all data are within a Github repository associated with this work (\url{}) that is also archived on Zenodo (INSERT URL AT TIME OF PROOFS). This repository also contains Python code related to calculations, visualizations and statistical tests discussed herein.

\bibliography{YZ_ref}

@@

% I conducted a comparative study of modified IZZI paleointensity experiments on both the diabase and anorthosite, with a group treated with low-field AF demagnetization after each in-field step and another group without. Paleointensity experiments for most diabase specimens result in non-ideal Arai plots often with poor pTRM checks, making them difficult to interpret for paleointensity estimate (Fig. \ref{fig:pmag}). These results can be interpreted to be associated with observed color change of the diabase after heating, which suggests alterations of the oxides. The experiments conducted on anorthosite samples, on the other hand, had a high success rate of $>$ 50\%. Moreover, the resulting success rate for the anorthosite group with the AF treatment is even higher. The more straight Arai plots are likely due to the AF steps mitigating the exhibition of pTRM tails often associated with non-ideal behavior of multi-domain grains that are demagnetized by the pre-treatment step. However, as the nearly pure anorthosite samples did not display dramatic alterations that are easily detectable with common petrographic techniques, a question rises as why some of the compositionally similar anorthosites passed the paleointensity selection results and some did not.

>>>>

\begin{abstract}
New high-quality absolute paleointensity data from \textit{ca.} 1092 Ma anorthosite xenoliths within the Beaver River diabase intrusions of the North American Midcontinent Rift record an average virtual axial dipole moment (VADM) estimate of $\sim$80 ZAm$^2$. This cooling rate corrected paleointensity estimate from Midcontinent Rift intrusions is similar to previous results from extrusive units that are dated to be between \textit{ca.} 1108 Ma and \textit{ca.} 1085 Ma. Rock magnetic analyses and magnetic imaging reveal that the anorthosite xenoliths can often have magnetic mineralogy dominated by grains with high coercivities, which can yield high success rates passing our selection criteria relative to their diabase hosts. Taken together, new and previously published paleointensity data from the Midcontinent Rift support that a strong geomagnetic field  persisted through 25 million years of the late Mesoproterozoic. These data are inconsistent with interpretations of continuous progressive decay of Earth's geodynamo strength through the Proterozoic Eon.

@@

The strong geodynamo today is collectively powered by thermal convection and compositional convection in the liquid outer core. Studies have found that Earth may have sustained a geodynamo through the Archean \cite{Tarduno2014a}, and possibly as far back as the Hadean \cite{Tarduno2020a}. However, recent estimates of the thermal conductivity of the core suggest that the nucleation of the inner core nucleation could be $<$500 Ma, more recent than previous thought. Withouth the power of compositional convection provided by light elements being rejected into the outer core from the inner core, any identification of an earlier geodynamo would indicate that the core was stirred by other power sources, such as precipitation of Mg \cite{ORourcke2016a}, or perhaps
that the dynamo was generated by exotic processes like a convecting basal magma ocean \cite{Ziegler2013a}. In any case, a geodynamo solely driven by thermal conduction through the core-mantle boundary (CMB) prior to the nucleation of the inner core is likely to decay persistently due to heat loss.

@@

Non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained from Midcontinent Rift rock units. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. \citeA{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas \citeA{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by \siteA{Sprain2018a} who applied stricter paleointensity selection criteria and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. Despite the low success rate of paleointensity experiments, specimens that yield dominantly single-slope paleointensity results from the \textit{ca.} 1105 Ma flows of the Osler Volcanic Group, ca. 1095 Ma flows of the Mamainse Point volcanics \cite{Sprain2018a}, as well as the \textit{ca.} 1085 Ma flows of the Lake Shore Traps \cite{Kulakov2013a} suggest VADM values similar to that of the present-day geomagnetic field \cite{Sprain2018a}. Taken together, these rock units that have been dated to span $\sim$25 myr during the development of the Midcontinent Rift magmatism \cite{Swanson-Hysell2019a} may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

Although it is typical of paleointensity experiments to suffer from low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested plagioclase crystals can protect magnetic minerals that exsolved within them against thermochemical alteration during paleointensity experiments. In this study, we present new paleointensity results from ``IZZI" absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the Beaver River anorthosite xenoliths has exceptionally high success rate. Thermal remanent magnetization (TRM) anisotropy and acquisition experiments reveal that the magnetite-dominated remanence of the anorthosite specimens has low TRM anisotropy and linear TRM acquisition. Magnetic imaging shows that the anorthosite specimens to have and dominant magnetic carriers interstitial to the plagioclase and are well shielded. With more single-slope, high quality paleointensity data, we seek to compare paleointensity records from the Midcontinent Rift intrusives with those from the extrusives; to test the hypothesis that the Earth's geomagnetic dipole strength remained at high values throughout the Midcontinent Rift; and to further enrich the reliable paleointensity records during the $\sim$25 myr of Midcontinent Rift magmatic activity.

@@

\begin{table}
\caption{Quality criteria used for accepting or rejecting paleointensity data. See the text for more detail.} 
\centering
\begin{tabular}{ccccccccc}
\hline
MAD (o) & Beta (per cent) & DANG (o) & FRAC& SCAT & GAP-MAX & NpTRM & N Arai\\
\hline
20 & 15 & 5 & 0.6 & TRUE & 0.25 & 2 & 4 \\
\hline

@@

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{MPMS.pdf}
\centering
\caption{\footnotesize{Low-temperature magnetic property measurement system (MPMS) experiment results. In the field-cooled (FC) experiments, the magnetization was measured upon warming following the specimen having cooled in an applied field of 2.5 T from 300 to 10 K. In the zero-field-cooled (ZFC) experiment, a low-temperature saturation isothermal remanence (LTSIRM) of 2.5 T was applied at 10 K after the specimen cooled in a (near-)zero field. In the room-temperature saturation isothermal remanence (RTSIRM) experiment, the sample was pulsed with a 2.5 T field at room temperature ($\sim$300 K) and then cooled to 10 K and warmed back to room temperature in a (near)zero field. Specimen AX6-1r is from anorthosite AX6 which passed our paleointensity selection. It has a well-defined Verwey transition $\sim$120 K. Specimens AX3-2r-ni and BD2-2r have prominent Verwey transition but the transition temperatures are suppressed below $\sim$120 K. Specimen BD2-2r has a consistently higher moment during the zero-field-cooled step than during the field-cooled step. This is consistent with the interpretation that multidomian magnetic carriers existed in significant quantity within this sepcimen.}}
\label{fig:MPMS}
\end{figure}

@@

anisotropy of full TRM experiment on accepted specs show low degree of bulk sample scale full TRM anisotropy, with a range of ---. Cooling rate correction based on \cite{Halgedahl1980a} suggest a lab over estimate of B ancient of 35 \%, assuming constant cooling rate in natural condition. The cooling rate corrected field is $\sim$40 $\mu$T which is similar to today's field strength at similar latitude.

\acknowledgments
Project research was supported by NSF CAREER grant EAR-1847277 to N.L.S.-H. and an Institute on Lake Superior Geology Student Research Fund grant to Y.Z. Permits for fieldwork and sampling from the Minnesota Department of Natural Resources are gratefully acknowledged. We thank James Pierce and Blake Hodgin for their assistance in the field. We thank Jim Miller for providing us with the Duluth Complex anorthosite thin section. We thank Dario Bilardello, Peat Solheid, Mike Jackson, Josh Feinberg, and Bruce Moskowitz for their tremendous help of instrumental operations, data interpretations, and research guidance. Y.Z. thanks the IRM for the U.S. Visiting Student Fellowship. Paleomagnetic data associated with this study are available within the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED) and all data are within a Github repository associated with this work (\url{}) that is also archived on Zenodo (INSERT URL AT TIME OF PROOFS). This repository also contains Python code related to calculations, visualizations and statistical tests discussed herein.  
\bibliography{YZ_ref}

@@
